Experiences Gamifying Developer Adoption\\ of Practices and Tools

### 2014-03-14 paper_id51.tex
Submitted on Thursday

\documentclass{sig-alternate}
% *** CITATION PACKAGES ***
%
\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

@@

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/
\usepackage{balance}
\usepackage{mdframed}
\usepackage{enumitem}
%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{ICSE}{'14, May 31 -- June 7, 2014, Hyderabad, India}
\CopyrightYear{14}
\crdata{978-1-4503-2768-8/14/05}
% --- End of Author Metadata ---

@@

As software development practices evolve, toolsmiths face the continuous challenge of getting developers to adopt new practices and tools.   We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.  We present results from a pre-study survey of 130 developers' opinions on gamification and motivation, usage data from a pilot with an intact team of six developers of a game on code navigation practices, and feedback collected in post-pilot interviews.  Our pre-study survey showed that most developers were interested in gamification, though some have strong negative opinions.  Pilot results show that two of the six pilot developers adjusted their practices when presented with competitive game elements.
%We found that a software engineering game with competitive elements influences some developers towards changing their practices.
%We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
%
% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).  
%
%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
%
% Embedding Blaze in the developer's work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.  
%
%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.  We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
\end{abstract}

@@

Sascha is a developer with years of experience maintaining code, who knows how to navigate through code fairly well.  Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE mainly out of habit.  When Sascha's company started a game around code navigation in Visual Studio, he became interested in being a part of the fun.  Sascha tried the code navigation tools and practices that were part of the game and soon found himself at the top of the game's leaderboard.  After participating in the game, Sascha used the new practices more often and felt recognized for his achievement.
%Deleted the following due to Review #1 comments who concluded the paper was about improving productivity using structured navigation. After learning the new practices Sascha spends less time navigating code to accomplish his maintenance tasks.

@@

Would software developers writing serious software be motivated by this gamification concept? To answer this question, this work contributes an assessment of developers' response to game-like elements tied to their use of new tools and practices.  We assess 130 developers' response and relevant concerns about the idea of gamification through a web-based survey.  We piloted Blaze, our tool that introduces game elements focused on improving navigation practices into the developer's IDE.  Developers' reactions to gamification were mostly positive with a few detractors, and results of piloting show that game elements influence some developers, but have no effect on others' use of tools and practices.

The rest of this paper is organized as follows:  Section 2 reviews related work, Section 3 reviews results from the pre-study survey, Section 4 discusses the design considerations for developing the Blaze tool/game, Section 5 discusses the design of the Blaze pilot study, Section 6 results from the pilot study and post-pilot survey, Section 7 covers threats to validity, and Section 8 has our conclusions.

Johnson and Kou defined Zorro \cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.  Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.  Using the distinct events developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario. In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy when classifying episodes into their proper TDD scenarios.  The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.  In our study, we attempt to present data collected directly to developers in order to influence them to change their practices.

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using PROM include a series of studies on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  Studies using the PROM tool correlate the time spent editing with code metrics obtained through source code analysis.  Applying these data to specific research questions, such as whether refactoring improves productivity \cite{Moser2008Case}, shows the utility of combining automated effort data with code metric data.  Taking a different focus on techniques, our study used more detailed events captured from the IDE to detect commands used to navigating through the code in addition to knowing duration of edit sessions for code modules.

Murphy-Hill et al. study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users\cite{MurphyHill2012Improving}.  Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history. Novice users were better served by recommendation algorithms that did not require a long history, while more expert users benefited from more sophisticated algorithms that included more history. The ``Most Widely Used'' algorithm recommends commands based on the collective usage profile of the team and performed nearly as well as the more sophisticated algorithms. The paper presents the methodology for making recommendations and evaluating their accuracy. Our work takes a further step evaluating how well tools can influence developers to use specific recommended commands.

When discussing gamification, concerns are raised about replacing an activity like software development, which is intrinsically motivated, with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, we reviewed several studies on motivation and gamification.

Hamari et al. analyze achievements patterns in game systems and provide guidance on defining successful achievement systems.  They recommend achievement awards be setup as a secondary form of scoring so players have multiple objectives available for a particular game.  Players should feel like they earned achievements in the game instead given trivial tasks.

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each other's rankings while the control group only sees their personal rankings.  In our study, we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.

@@

Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?  We segmented results by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicate 75-88\% of developers in India, Poland, and Finland were at least somewhat interested.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.

Comments from participants reflected a more interesting controversial reaction than these measurements might otherwise indicate.  A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams.'' indicating they are under considerable pressured to get the work done.  Another developer reacted with a statement that identifies the kind of intrinsic motivation we hope for ``I am using practices and tools because I want to do my job well',' indicating that gamification was unnecessary to motivate people to improve.

@@

Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.  Our results land on both sides of the 75\% mark reported by the Saatchi and Saatchi study.

@@

Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  The question provided a graduated scale for the scope of who data was shared with and divided the questions between whether sharing was anonymous or not. Figure \ref{fig:comfortwithsharing}  shows, as the community with which the data is shared gets smaller, more developer are comfortable with sharing the data.  If we take scope as the main concept, and limit the sharing to ``My Team'', we can say that 90\% of respondents are comfortable with sharing either anonymous or non-anonymous data if the scope of data were limited to ``My Team''.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data was not kept anonymous.  
The first conclusion from these questions on sharing data is that people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

@@

Just behind team competition is management mandate where developers would be asked to follow specific practices by the organization.  The lowest was having badges posted on their social network profile where 32\% ranked this in 1st or 2nd place.   Badges and management mandate received the most negative written comments in the survey.  A developer from the United States captured the lower opinion of badges that several participants expressed ``I am motivated because ABB allows me to choose the tools that best work for me. A badge will not really do anything for me.''  We conclude from this result, supported by findings in \cite{MurphyHill2012Improving,wbsnipes:Hill2011Peer}, that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.

@@

Following the pre-study findings, we implemented a game-like tool for developers to use in Visual Studio.  The tool focused on motivating the adoption of good practices by including elements of games in the development environment.  The following sections describe how we selected a practice suitable for gamification, designed the game, and included competitive elements in the game.

@@

We defined the first criteria for selecting a practice based on prior work  applying gamification to other domains outside software development. 
In their paper on game design patterns, Hamari et al. issued a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework}.  Thus we avoided selecting practices like bug fixing or task completion because they are simply requirements of the job.

Another criteria imposed by the instrument limits us to practices performed entirely in the IDE so they can be monitored using Blaze.  We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM (configuration management) submissions.

Frequent CM submission and static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.  Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro \cite{V:Johnson2007Automated} thus may provide a candidate for future gamification application.   
%reference for debugging?
Debugging practices have developer community opinions on optimal methods, however, detailed debug practices with modern IDEs have not received research community attention, thus are also reserved for future work.

Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} (discussed in Related Work \ref{sec:RelatedWork}) identify structured navigation as a practice more effective developers use when maintaining programs.  Structured navigation fits the requirements of a practice contained in the IDE, has proven benefits, and is not specifically assigned task for a developer.  Thus we selected structured navigation as the practice to demonstrate gamification.

@@

By selecting an intact team for the pilot, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition between teammates.    The leaderboard allows developers to see their own score in relation to the top five people on the team.  However, the other participants' identifiers are auto-generated as three capital letters assigned based on their position on the leaderboard.  Not knowing who of your colleagues was ahead of you could reduce motivation, but this avoids one potential source of conflict in an intact team.

@@

Utilizing the continuous monitoring capabilities of Blaze, we designed the pilot to conduct a longitudinal study of the effect of Blaze on developers' navigation practices.  The Blaze tool revealed information about navigation in three stages of one week duration each.  A post-pilot survey gathered developers' opinions of Blaze and some confounding factors that could influence results.

@@

Part of the research objective was to determine whether we can drive adoption of a practice through gamification.  As previously discussed, we designed the game in Blaze to influence developers towards using more structured navigation practices.  We constructed the interventions during the pilot to roll out in stages each week they use the tool.  This allowed us to establish a baseline and determine the changes as the pilot continued.

@@

After a week's data collection marking the end of the first stage, Blaze started to automatically pop-up a window containing a button for a web page with information.  The page included a link to download Sando  and information about using Sando in a video demo.  The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.  The ``About Blaze'' page contained details of the point scoring system and background on Blaze's purpose.  The usefulness of the communication site is rated as part of our post-pilot  survey.

In the third stage, developers received instant feedback from Blaze on the points they accumulated during the development session.  The tool's appearance at this stage is shown in Figure \ref{fig:blazeWindow}.  The ``Info'' button provides a level indicator via colored contents and activates a chart display window when clicked.  Points shows the points accumulated for the developer's history.  Level is the level of the game achieved based on an exponentially growing curve.   The developer could click on the Leaderboard tab to see how their scores compared with the others in the pilot.  This final configuration remained as the display until the end of the pilot.

@@

\subsection{Pilot Participants}

@@

We  assured the team about the confidentiality of the data, and the management reassured them that the data collected by the research group will not be used for any other purpose other than the study.  Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study. Their names and data were kept confidential to ensure that no one apart from the research team knows he/she is participating in the pilot.  To distinguish individuals in the data each user is assigned a generated unique ID.  Developers could chose to share the unique ID with us or keep it confidential.  All were willing to share their unique ID with us during the follow-up feedback sessions.

Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.  Data at this level of detail has not been collected in ABB development organizations.  Researchers were careful to create the system where developers controlled whether we could associate their identity with their data, and fortunately management bought into the necessity for participant anonymity.

\subsection{Pilot Survey}

The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers' navigation practices and knowledge.  The survey questions were segmented into portions that evaluate prior and post knowledge of navigation practices, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.

To find out whether Blaze increased developers'  knowledge of structured navigation commands, we asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale.  Then we asked them to contrast their knowledge after the pilot to assess what they learned from the pilot.

The other key question is did Blaze have a perceived effect on their navigation practices.  Here too we asked them to rate how much they used structured navigation prior to the pilot using a 4 choice rating of ``not at all'', ``some'', ``many times'', and ``as much as possible''.  Then we asked whether they used structured navigation commands ``about the same amount'', ``more'',  or ``a lot more'' during the pilot.  These two questions assessed their perceptions of structured navigation command use.  We compared the answers with the usage data collected from Blaze.

@@

\subsection{Pilot Survey Results}

We conducted an in-person post pilot survey in order to cross check the quantitative data from Blaze with developers' perception of their practices and their demographic information.

The pilot survey queried developers' relevant demographic data to help understand some differences in results.  We asked developers to provide their years of experience so we could evaluate any effects experience might have.  The developers in the study had between 7 and 14 years experience.  One developer who dropped out of the study said their lower level of experience compared to their peers influenced their decision to stop participating.

@@

\subsection{Pilot Data Observations}

@@

In another processing step, Blaze aggregates events along the time-line into sessions.  A session is an abstract concept grouping multiple events into a sequence with a beginning and an end.  Of the many possible ways to define sessions, we chose to define sequences to contain all the events that transpire before and during a set of edits to a file.  When the developer edits a file different from the file last edited, all the events after the last edit to the previous file are included in the new sequence for the new file.  For example, the developer edits file A, then navigates through files B and C finally landing at line 100 of file D where they make another edit.  All the events between the edit of file A and the edit of file D are considered part of the edit session ending with the edit of file D.  When this edit is followed by multiple edits to file D, the additional edits are included in the same sequence ending with the edits to file D.  In other words, a sequence is defined to contain the set of actions that occur between  Although not at the method level, this way of generating sequences is similar to how Robillard et al. generated sequences based on located methods for their study \cite{wbsnipes:Robillard2004How}.

\subsubsection{Observations About Drivers of Edit Session Duration}
\label{sec:driversOfEditSession}
We selected navigation as the subject of the pilot game because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    In this analysis, we ask whether we are focusing on the right area of developers' activity in Visual Studio.

In order to test whether navigation is an important area to address, we construct a hypothesis that using unstructured navigation is an important factor in developer productivity.  Edit session duration, we assert by opinion, is a factor related to developer productivity though there are many other factors we could consider.  With the data available from Blaze, we can explore the categories of events correlated with the duration of edit sessions over the period of the pilot.

>>>>

\documentclass{sig-alternate}
\pdfpagewidth=8.5in
          \pdfpageheight=11in
% *** CITATION PACKAGES ***
%
\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

@@

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/
\usepackage{balance}
\usepackage{mdframed}
\usepackage{enumitem}
\usepackage{url}
%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools

\begin{document}
%ICSE '14, May 31 - June 07 2014, Hyderabad, India

% --- Author Metadata here ---
\conferenceinfo{ICSE}{'14, May 31 -- June 7, 2014, Hyderabad, India}
\CopyrightYear{is held by the owner/author(s). Publication rights licensed to ACM.}
\crdata{978-1-4503-2768-8/14/06}

% --- End of Author Metadata ---

@@

As software development practices evolve, toolsmiths face the continuous challenge of getting developers to adopt new practices and tools.   We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.  We present results from a pre-study survey of 130 developers' opinions on gamification and motivation, usage data from a study with an intact team of six developers of a game on code navigation practices, and feedback collected in post-study interviews.  Our pre-study survey showed that most developers were interested in gamification, though some have strong negative opinions.  Study results show that two of the six study developers adjusted their practices when presented with competitive game elements.
%We found that a software engineering game with competitive elements influences some developers towards changing their practices.
%We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
%
% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).  
%
%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
%
% Embedding Blaze in the developer's work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.  
%
%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.  We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
\end{abstract}

@@

Sascha is a developer with years of experience maintaining code, who knows how to navigate through code fairly well.  Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio Integrated Development Environment (IDE) mainly out of habit.  When Sascha's company started a game around code navigation in Visual Studio, he became interested in being a part of the fun.  Sascha tried the code navigation tools and practices that were part of the game and soon found himself at the top of the game's leaderboard.  After participating in the game, Sascha used the new practices more often and felt recognized for his achievement.
%Deleted the following due to Review #1 comments who concluded the paper was about improving productivity using structured navigation. After learning the new practices Sascha spends less time navigating code to accomplish his maintenance tasks.

@@

Would software developers writing serious software be motivated by this gamification concept? To answer this question, this work contributes an assessment of developers' response to game-like elements tied to their use of new tools and practices.  We assess 130 developers' response and relevant concerns about the idea of gamification through a web-based survey.  As a research study, we piloted the Blaze tool that introduces game elements focused on improving navigation practices to developers using Visual Studio.  Developers' reactions to gamification were mostly positive with a few detractors, and results of the study show that game elements influence some developers, but have no effect on others' use of tools and practices.

The rest of this paper is organized as follows:  Section 2 reviews results from the pre-study survey validating the acceptance of gamification with developers, Section 3 contains the design considerations for developing the Blaze tool/game, Section 4 gives the study design for attempting to influence developers` work patterns with Blaze, Section 5 results from the study and post-study survey, Section 6 covers related work, Section 7 discusses threats to validity, and Section 8 has our conclusions.

@@

A consulting firm, Saatchi and Saatchi, use a question in a survey that is often cited in gamification initiatives.  This study reuses this question \cite{wbsnipes:SaatchiGameification} tailoring it to software development by asking ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?  Results are segmented by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicate 74\% of developers in India, Poland, and Finland were at least somewhat interested, while the US, Sweden, and Switzerland had slightly lower interest in applying gamification to their work.

Comments from participants reflected a more interesting controversial reaction than these measurements might otherwise indicate.  A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a general concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams.'' indicating they are under considerable pressured to get the work done.  Another developer reacted with a statement that expresses a positive intrinsic motivation ``I am using practices and tools because I want to do my job well',' indicating that gamification was unnecessary to motivate people to improve.

@@

The Figure \ref{fig:gamification} compares the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.  Our results land on both sides of the 75\% mark reported by the Saatchi and Saatchi study.

@@

Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  The question provided a graduated scale for the scope of who data was shared with and divided the questions between whether sharing was anonymous or not. Figure \ref{fig:comfortwithsharing}  shows, as the community with which the data is shared gets smaller, more developer are comfortable with sharing the data.  If we take scope as the main concept, and limit the sharing to ``My Team'', we can say that 90\% of respondents are comfortable with sharing either anonymous or non-anonymous data if the scope of data were limited to ``My Team''.  Developers were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data was not kept anonymous.  
The first conclusion from these questions on sharing data is that people are very willing to share information that could help their team. Second, sharing within the company is acceptable with anonymized data.

@@

Just behind team competition is management mandate where developers would be asked to follow specific practices by the organization.  The lowest was having badges posted on their social network profile where 32\% ranked this in 1st or 2nd place.   Badges and management mandate received the most negative written comments in the survey.  A developer from the United States captured the lower opinion of badges that several participants expressed ``I am motivated because ABB allows me to choose the tools that best work for me. A badge will not really do anything for me.''

 The respones on this question, supported by findings in \cite{MurphyHill2012Improving,wbsnipes:Hill2011Peer,Dubois2013Understanding}, influenced the development of the studied game to include a competitive leaderboard and a common recommendation contest to influence ABB developers to change their practices.

@@

Learning from the pre-study findings, we implemented a game-like tool for developers to use in Visual Studio.  The tool focused on motivating the adoption of good practices by including elements of games in the development environment.  This section describes the selection of a practice suitable for gamification, considerations for designing the game, and competitive elements in the game.

@@

We defined the first criteria for selecting a practice based on prior work  applying gamification to other domains outside software development. 
In their paper on game design patterns, Hamari et al. stated that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework}.  Thus we avoided selecting practices like bug fixing or task completion because they are simply requirements of the job.

Another criteria imposed by the instrument limits us to practices performed entirely in the IDE so they can be monitored using Blaze.  Potential practices for gamification include  test-driven development, refactoring tool use, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent configuration management submissions.

Frequent configuration management submission and static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.  Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro \cite{V:Johnson2007Automated} thus may provide a candidate for future gamification application.   Refactoring tool use also has studies that make it a good candidate for gamification \cite{V:MurphyHill2012How,Moser2008Case}.

Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} (discussed in Related Work \ref{sec:RelatedWork}) identify structured navigation as a practice more effective developers use when maintaining programs.  
Structured navigation is the use of navigation commands and tools that follow or represent the structure of the code.  Using structured navigation provides a shortcut that confirms the existence of and affirms the mental model the developer has of the code.  
Robillard et al.`s  experiment~\cite{wbsnipes:Robillard2004How} showed developers who maintain mental models and navigate structurally are more effective than other developers when given the same maintenance task and code artifacts.
Structured navigation fits the requirements of a practice contained in the IDE, shows benefits, and is not specifically assigned task for a developer.  Thus we selected structured navigation as the practice to demonstrate gamification.

@@

By selecting an intact team for the study, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition between teammates.    The leaderboard allows developers to see their own score in relation to the top five people on the team.  However, the other participants' identifiers are auto-generated as three capital letters assigned based on their position on the leaderboard.  Not knowing who of your colleagues was ahead of you could reduce motivation, but this avoids one potential source of conflict in an intact team.

@@

Utilizing the continuous monitoring capabilities of Blaze, we designed the study to conduct a longitudinal study of the effect of Blaze on developers' navigation practices.  The Blaze tool revealed information about navigation in three stages of one week duration each.  A post-study survey gathered developers' opinions of Blaze and some confounding factors that could influence results.

@@

The key research objective was to determine whether we can drive adoption of a practice through gamification.  As previously discussed, we designed the game in Blaze to influence developers towards using more structured navigation practices.  We constructed the interventions during the study to roll out in stages each week they use the tool.  This allowed us to establish a baseline and determine the changes as the study continued.

@@

After a week's data collection marking the end of the first stage, Blaze started to automatically pop-up a window containing a button for a web page with information.  The page included a link to download Sando  and information about using Sando in a video demo.  The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.  The ``About Blaze'' page contained details of the point scoring system and background on Blaze's purpose.  The usefulness of the communication site is rated as part of our post-study  survey.

In the third stage, developers received instant feedback from Blaze on the points they accumulated during the development session.  The tool's appearance at this stage is shown in Figure \ref{fig:blazeWindow}.  The ``Info'' button provides a level indicator via colored contents and activates a chart display window when clicked.  Points shows the points accumulated for the developer's history.  Level is the level of the game achieved based on an exponentially growing curve.   The developer could click on the Leaderboard tab to see how their scores compared with the others in the study.  This final configuration remained as the display until the end of the study.

@@

\subsection{Study Participants}

@@

We  assured the team about the confidentiality of the data, and the management reassured them that the data collected by the research group will not be used for any other purpose other than the study.  Developers were asked to volunteer for the study by their management, however, managers did not know who participated in the study. Their names and data were kept confidential to ensure that no one apart from the research team knows he/she is participating in the study.  To distinguish individuals in the data each user is assigned a generated unique ID.  Developers could chose to share the unique ID with us or keep it confidential.  All were willing to share their unique ID with us during the follow-up feedback sessions.

Management wished to have no knowledge of who was participating and took great care to remain detached from the study proceedings.  Data at this level of detail has not been collected in ABB development organizations.  Researchers were careful to create the system where developers controlled whether we could associate their identity with their data, and fortunately management bought into the necessity for participant anonymity.

\subsection{Study Survey}

The goal for the study follow-up survey was to collect feedback on the tool and the effect it had on the developers' navigation practices and knowledge.  The survey questions were segmented into portions that evaluate prior and post knowledge of navigation practices, key steps in the study itself, the influence of game elements on their navigation practices, typical tasks and demographic information.

To find out whether Blaze increased developers'  knowledge of structured navigation commands, we asked them to rate their knowledge of structured navigation commands prior to the study on a 1-10 scale.  Then we asked them to contrast their knowledge after the study to assess what they learned.

The other key question is did Blaze have a perceived effect on their navigation practices.  Here too we asked them to rate how much they used structured navigation prior to the study using a 4 choice rating of ``not at all'', ``some'', ``many times'', and ``as much as possible''.  Then we asked whether they used structured navigation commands ``about the same amount'', ``more'',  or ``a lot more'' during the study.  These two questions assessed their perceptions of structured navigation command use.  We compared the answers with the usage data collected from Blaze.

@@

\subsection{Study Survey Results}

We conducted an in-person post study survey in order to cross check the quantitative data from Blaze with developers' perception of their practices and their demographic information.

The study survey queried developers' relevant demographic data to help understand some differences in results.  We asked developers to provide their years of experience so we could evaluate any effects experience might have.  The developers in the study had between 7 and 14 years experience.  One developer who dropped out of the study said their lower level of experience compared to their peers influenced their decision to stop participating.

@@

\subsection{Study Data Observations}

@@

In another processing step, Blaze aggregates events along the time-line into sessions.  A session, similar to episodes defined by ~\cite{Kou2010Operational}, is an abstract concept grouping multiple events into a sequence with a beginning and an end.  Of the many possible ways to define sessions, we chose to define edit sessions to contain all the events that transpire between edits of different files.  When the developer edits a file different from the file last edited, all the events after the last edit to the previous file are included in the new session for the new file.  For example, the developer edits file A, then navigates through files B and C finally landing at line 100 of file D where they make another edit.  All the events between the edit of file A and the edit of file D are considered part of the edit session ending with the edit of file D.  When this edit is followed by multiple edits to file D, the additional edits are included in the same session ending with the edits to file D.   Events transpiring between edits may involve activities not considered as edit related such as building and testing.  This session definition allows us to quantify how much time in each category of events elapses before the developer edits another area of code.  Although not at the method level, this way of generating sessions is similar to how Robillard et al. generated sessions based on located methods for their study \cite{wbsnipes:Robillard2004How}.

\subsubsection{Observations About Drivers of Edit Session Duration}
\label{sec:driversOfEditSession}
We selected navigation as the subject of the Blaze study because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    In this analysis, we ask whether we are focusing on the right area of developers' activity in Visual Studio.

In order to test whether navigation is an important area to address, we construct a hypothesis that using unstructured navigation is an important factor in developer productivity.  Edit session duration, we assert by opinion, is a factor related to developer productivity though there are many other factors we could consider.  With the data available from Blaze, we can explore the categories of events correlated with the duration of edit sessions over the study period.

Using Weka~\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.  The attribute search method we chose was Greedy Step-wise search, which selects attributes based on their correlation with the target, and stops adding attributes when the next attribute decreases the correlation.    Defining the attributes as the sum of developer time spent by category in each edit session, the top 5 out of the 15 possible categories ranked by this process are as follows:
\begin{enumerate}[itemsep=0mm]
\item Unstructured navigation
\item Debug - events from running the debugger
\item Edit - events that modify the code
\item Other Actions - other events in visual studio that we do not sub-classify
\item Build - events from running a build
\end{enumerate}
